The state learner \\ {-- a super learner for right-censored data}

### 2024-04-29 tex/jrss-version/state-learner-jrss-style.tex

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone tools and in
  applications of causal inference to estimate nuisance parameters. The super
  learner is a machine learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. Unfortunately, the
  commonly used partial likelihood loss is not suited for super learning, and
  inverse probability of censoring weighted loss functions require a
  pre-specified estimator of the censoring distribution. To relax this, we
  introduce the state learner, a new super learner for survival analysis, which
  evaluates the loss based on the observed data simultaneously using libraries
  of predictions models for the event(s) of interest and the censoring
  distribution. We establish an oracle inequality for the state learner and
  investigate its performance through numerical experiments. We illustrate how
  the state learner allows us to estimate causal effects in a competing risks
  setting without having to pre-specify models for neither the cause-specific
  hazards of interest nor the censoring distribution.
\end{abstract}

@@

A super learner is a machine learning algorithm that combines a finite
set of learners into a meta learner by estimating prediction
performance in hold-out samples using a pre-specified loss function
\citep{van2007super}. When the aim is to make a prediction model,
super learners typically combine strong learners, such as Cox
regression models and random survival forests
\citep{gerds2021medical}.
% For targeted learning of a low dimensional
% target parameter in presence of a high-dimensional nuisance parameter,
% a super learner will usually also include the highly adaptive lasso
% \citep{van2011targeted, benkeser2016highly,van2017generally}.
While the general idea of combining strong learners based on cross-validation
data stems from earlier work \citep{wolpert1992stacked,breiman1996stacked}, the
name super learner is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed to perform
almost as well as the model which minimises the expected performance, i.e., the
model we would select if we could evaluate the prediction performance in an
infinite hold-out sample.

We are concerned with the choice of the loss function for super learning in
survival analysis. Existing super learner algorithms for right-censored data use
partial log-likelihood loss or inverse probability of censoring weighted loss
\citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,westling2021inference}.
The use of the partial log-likelihood loss restricts the class of learners and
excludes for example simple Kaplan-Meier based learners and also more complex
random survival forest algorithms. For this reason \cite{golmakani2020super}
restrict their learners to Cox proportional hazard models. A lesser known fact
is that a super learner constructed with the negative partial log-likelihood
loss implicitly depends on the censoring distribution
\citep{hjort1992inference,whitney2019comment}. A disadvantage of inverse
probability of censoring weighted loss functions is that they require a
pre-specified model for the censoring distribution. \cite{westling2021inference}
tackle this challenge by iterating between super learning of the censoring
distribution and the event time distribution.

In this article we define the state learner, a new super learner for
right-censored data, which simultaneously evaluates the loss for learners of the
event time distribution and the censoring distribution. The loss function which
is used to define the state learner is only based on observable
quantities. The state learner can be applied to all types of survival
estimators, works in the presence of competing risks, and does not require a
single pre-specified estimator of the conditional censoring distribution. To
analyze the theoretical properties of the state learner we focus on the
so-called discrete super learner which `combines' the library of
learners by picking the one that minimises the cross-validated loss. The state
learner uses separate libraries to model each competing event and the censoring
distribution. We show that the oracle selector of the state learner is
consistent if all libraries contain a consistent learner and prove a finite
sample oracle inequality.

The state learner can be used to select a model which predicts the probability
of an event based on covariates in the presence of competing risks. Another
application is in targeted learning where conditional event probabilities occur
as high-dimensional nuisance parameters which need to be estimated at a certain
rate \citep{van2011targeted, rytgaard2021estimation, rytgaard2022targeted}. We
show how a targeted estimator can be obtained from the state learner, and that a
second order product structure for the asymptotic bias term of the targeted
estimator is retained when the state learner is used to estimate nuisance
parameters.

The article is organised as follows. We introduce our notation and framework in
Section~\ref{sec:framework}. In Section~\ref{sec:super-learning} we define super
learning in general with right-censored data, and in
Section~\ref{sec:relat-liter-exist} we review existing methods.
Section~\ref{sec:super-learner-simple} introduces our novel proposal, the state
learner, and Section~\ref{sec:theor-results-prop} provides theoretical
guarantees for the state learner. In Section~\ref{sec:targeted-learning} we
discuss the use of the state learner in the context of targeted learning. We
report some numerical experiments in Section~\ref{sec:numer-exper} and analyze a
prostate cancer data set in Section~\ref{sec:real-data-appl}.
Section~\ref{sec:discussion} contains a discussion of the merits and limitations
of our proposal. Appendices~\ref{sec:proof-proposition}
and~\ref{sec:state-learner-with} contain proofs.

@@

In a competing risk framework \citep{andersen2012statistical}, let \( T\) be a
time to event variable, \(D\in\{1,2\}\) the cause of the event, and
$X \in \mathcal{X}$ a vector of baseline covariates taking values in a bounded
subset \( \mathcal{X} \subset \R^p \), \( p\in\N \). Let $\tau< \infty$ be the
maximal length of follow-up. We use \( \mathcal{Q} \) to denote the collection
of all probability measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
such that \( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard functions are
defined by
\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \) such that
\begin{equation*}
  % \label{eq:cum-haz}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*}
For ease of notation we assume throughout that \( \Lambda_j(\blank \mid x) \) is
continuous for all \( x \) and \( j \). We denote by \(S\) the conditional
event-free survival function,
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
Let \( \mathcal{M} \) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterised by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)&= \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=1\}}}\\
                        &\quad\left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

We consider the usual right-censored setting in which we observe data
\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability measures
on the sample space
\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such that
\(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume that the event
time and the censoring time are conditionally independent given covariates,
\( T \independent C \mid X \). This implies that any distribution
\( P \in \mathcal{P} \) is characterised by a distribution
\( Q \in \mathcal{Q} \) and a conditional cumulative hazard function for \( C \)
given \( X \) \citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard function for
censoring. We assume that \( \Gamma(\blank \mid x) \) is continuous for all
\( x \), and let \(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
survival function of the conditional censoring distribution. In our setting with
competing risks, this yields
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) &= \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
                           &=\left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M} \). We
also have
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*}
We further assume that there exists \(\kappa<\infty\) such that
\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
Under these assumptions, the conditional cumulative hazard functions
\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M\).

\section{Super learning with right-censored survival data}
\label{sec:super-learning}

A super learner estimates a parameter $\Psi$ which can be identified from the
observed data distribution \(P\in\mathcal P\). In this section, to introduce the
concept of super learning, we simply consider estimation of the function
\( \Lambda_{j}\). In this case, the parameter \(\Psi:\mathcal P\to\mathcal M\)
is identified via equation \eqref{eq:lambdaj} by \(\Psi(P)=\Lambda_j\).

As input to the super learner we need a sample \( \data_n=\{O_i\}_{i=1}^n \) of
i.i.d.\ observations from some unknown \( P \in \mathcal{P} \) and a finite
collection of candidate learners $\mathcal{A}$. Each learner
\(a \in \mathcal{A}\) is a map
\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}\) which takes a data set as
input and returns an estimate $a(\data_n) \in \mathcal{M}$ of $\Lambda_{j}$.
% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
In what follows, we use the short-hand notation
\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
performance of \(a \in \mathcal{A}\) using a loss function
\(L\colon \mathcal{M} \times \mathcal{O} \rightarrow \R_+\) by
estimating the expected loss \(P[L(a(\data_n), \blank)]\) using
cross-validation. Specifically, the expected loss of $a\in\mathcal A$
is estimated by splitting the data set $\data_n$ into $K$ disjoint
approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  % \label{eq:cv-risk-est}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
The oracle learner is defined as the learner that minimises the
average loss according to the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner depend on the
library of learners and on the number of folds \(K\), and that the oracle
learner is a function of the data and the unknown data-generating distribution.
These dependencies are suppressed in the notation.

@@

Machine learning based on right-censored data commonly use the negative partial
log-likelihood as loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
However, this loss function is unsuited for super learning, because many
canonical survival learners (e.g., the Kaplan-Meier estimator, random survival
forest, and semi-parametric Cox models) provide cumulative hazard functions that
are piece-wise constant in the time argument, and hence assign zero probability
to event times not observed in the training data. This implies that when data
are observed in continuous time, any of these learners will almost surely have
infinite loss in any independent hold-out sample according to the negative
partial log-likelihood loss.

When a proportional hazards model is assumed, the baseline hazard function can
be profiled out of the likelihood to give a new partial log-likelihood loss
\citep{cox1972regression}, which has been suggested as a loss function for super
learning \citep{golmakani2020super,verweij1993cross}. While this allows the
library of learners to include Cox' proportional hazard models, the drawback is
that the library is now restricted to include only these models.

@@

As far as we are aware, the only existing attempt at avoiding the need to
pre-specify a censoring model is a recent proposal suggested independently by
\cite{han2021inverse} and \cite{westling2021inference}. The authors do not
consider competing risks but suggest to iterate between learning \( \Lambda \)
and $\Gamma$ using IPCW loss functions and select the final learner when the
iterative procedure has converged. To the best of our knowledege, no general
theoretical guarantees exist for this procedure. It seems unclear whether this
iterative procedure will always be able to find a correctly specified model in a
given library.

@@

The state learner requires three libraries of learners, \(\mathcal{A}_1\),
\( \mathcal{A}_2 \), and \( \mathcal{B} \), where \(\mathcal{A}_1\) and
\( \mathcal{A}_2\) contain learners of the conditional cause-specific cumulative
hazard functions of the event time distribution \(\Lambda_1\) and
\( \Lambda_2\), respectively, and \(\mathcal{B}\) contains learners of the
conditional cumulative hazard function of the censoring
distribution. % We further

@@

$\mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})$ of learners

@@

As described in Section~\ref{sec:super-learning}, each learner

@@

\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \) is evaluated using
the cross-validated loss,

@@

and the discrete state learner is

>>>>

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone
  tools and in applications of causal inference to estimate nuisance
  parameters. The super learner is a machine learning algorithm which
  combines a library of prediction models into a meta learner based on
  cross-validated loss. In right censored data, loss needs to be estimated somehow.
  % based on a model for the censoring distribution.
  % Unfortunately, the
  % commonly used partial likelihood loss is not suited for super learning, and
  % inverse probability of censoring weighted loss functions require a
  % pre-specified estimator of the censoring distribution. To relax this,
  We introduce the state learner, a new super learner for survival analysis, which
  simultaneously evaluates libraries
  of prediction models for the event of interest and the censoring
  distribution. The state learner can
be applied to all types of survival models, works in the presence
of competing risks, and does not require a single pre-specified
estimator of the conditional censoring distribution.
  We establish an oracle inequality for the state learner and
  investigate its performance through numerical experiments. We illustrate how
  the state learner allows us to estimate causal effects in a competing risks
  setting without having to pre-specify models for neither the cause-specific
  hazards of interest nor the censoring distribution.
\end{abstract}

@@

Medical decision making is traditionally based on the discrepancy
between intuitive jugdment and the law of chance
\citep{redelmeier1995probability}. We are concerned with the
statistical models that provide probabilistic predictions about the
medical future of a patient. We have two different applications of
these models in mind. One is as a stand-alone tool where the model
provides information for a freshly diagnosed patient. The other is as
input to an algorithm which estimates a causal target parameter such
as an average treatment effect. The models are learned on data from
previous patients. A super learner is a machine learning algorithm
that combines a finite set of learners into a meta learner by
estimating prediction performance in hold-out samples using a
pre-specified loss function \citep{van2007super}. When the aim is to
make a prediction model, super learners combine strong learners, such
as Cox regression models and random survival forests
\citep{gerds2021medical}.
% For targeted learning of a low dimensional
% target parameter in presence of a high-dimensional nuisance parameter,
% a super learner will usually also include the highly adaptive lasso
% \citep{van2011targeted, benkeser2016highly,van2017generally}.

While the general idea of combining strong learners based on
cross-validation data stems from earlier work
\citep{wolpert1992stacked,breiman1996stacked}, the name super learner
is justified by an oracle inequality
\citep{van2003unicv,vaart2006oracle}: The super learner is guaranteed
to perform almost as well as the model which minimises the expected
performance, i.e., the model we would select if we could evaluate the
prediction performance in an infinite hold-out sample. The choice of
the loss function for super learning in survival analysis deserves a
closer look. Existing super learner algorithms for right-censored data
use partial log-likelihood loss or inverse probability of censoring
weighted loss
\citep{polley2011-sl-cens,keles2004asymptotically,golmakani2020super,westling2021inference}.
The use of the partial log-likelihood loss, however, restricts the
class of learners and excludes for example simple Kaplan-Meier based
learners and also more complex random survival forest algorithms. For
this reason \cite{golmakani2020super} restrict their learners to Cox
proportional hazard models. A lesser known fact is that a super
learner constructed with the negative partial log-likelihood loss
implicitly depends on the censoring distribution
\citep{hjort1992inference,whitney2019comment}. Inverse probability of
censoring weighted loss has been studied for decades
\citep{graf1999assessment, hothorn2006survival, gerds2006consistent}. A disadvantage of inverse probability of
censoring weighted loss estimators is that they require a
pre-specified model for the censoring
distribution. In \cite{westling2021inference} this problem is tackled by
iterating between super learning of the censoring distribution and the
event time
distribution. % ANDERS: mangler afrundning efter sidste saetning. eller maaske kan saetningen flyttes laenger ned?

In this article we define the state learner, a new super learner for
right-censored data, which simultaneously estimates the expected loss
of learners of the event time distribution and the censoring
distribution. The loss function which is used to define the state
learner is based on the observable
quantities. % ANDERS: hvad mener vi med observable quantities? distribution of the observed random variables?
The state learner can be applied to all types of survival models,
works in the presence of competing risks, and does not require a
single pre-specified estimator of the conditional censoring
distribution.  To analyze the theoretical properties of the state
learner we focus on the discrete super learner which combines the
library of learners by picking the one that minimises the
cross-validated loss
\citep{van2011targeted}. % ANDERS: er der bedre referencer fra Mark paper/bog for discrete super learner?
In the presence of competing risks, our algorithm uses separate
libraries to model the hazard rates of the competing risks and the
censoring distribution. We show that the oracle selector of the state
learner is consistent if all libraries contain a consistent learner
and prove a finite sample oracle inequality. The result of the state
learner algorithm is a model which predicts the probability of an
event based on covariates in the presence of competing risks
\citep{gerds2021medical}. The other application is in targeted
learning where conditional event probabilities occur as
high-dimensional nuisance parameters which need to be estimated at a
certain rate \citep{van2011targeted, rytgaard2021estimation,
  rytgaard2022targeted}. We show for a targeted estimator, which uses
the state learner to estimate nuisance parameters, that a second order
product structure holds for the asymptotic bias term.

The article is organised as follows. We introduce our notation and
framework in Section~\ref{sec:framework}. In
Section~\ref{sec:super-learning} we define general super learning
for right-censored data, and in Section~\ref{sec:relat-liter-exist}
we review existing methods.  Section~\ref{sec:super-learner-simple}
introduces our novel proposal, the state learner, and
Section~\ref{sec:theor-results-prop} provides theoretical guarantees
for the state learner. In Section~\ref{sec:targeted-learning} we
discuss the use of the state learner in the context of targeted
learning. We report results of our numerical experiments in
Section~\ref{sec:numer-exper} and analyze a prostate cancer data set
in Section~\ref{sec:real-data-appl}.  Section~\ref{sec:discussion}
contains a discussion of the merits and limitations of our
proposal. Appendices~\ref{sec:proof-proposition}
and~\ref{sec:state-learner-with} contain proofs.

@@

In a competing risk framework \citep{andersen2012statistical}, let \( T\) be a
time to event variable, \(D\in\{1,2\}\) the cause of the event, and
$X \in \mathcal{X}$ a vector of baseline covariates taking values in a bounded
subset \( \mathcal{X} \subset \R^p \), \( p\in\N \). Let $\tau< \infty$ be the
maximal length of follow-up. We use \( \mathcal{Q} \) to denote the collection
of all probability measures on \( [0,\tau] \times \{1,2\}\times \mathcal{X} \)
such that \( (T, D, X) \sim Q \) for some unknown \( Q \in \mathcal{Q} \). For
\(j\in\{1,2\}\), the cause-specific conditional cumulative hazard functions are
defined by
\( \Lambda_{j} \colon [0, \tau] \times \mathcal{X} \rightarrow \R_+ \) such that
\begin{equation*}
  % \label{eq:cum-haz}
  \Lambda_{j}(t \mid x) = \int_0^t\frac{  Q(T \in \diff s, D=j \mid X=x )}{Q(T \geq s \mid X=x )}.
\end{equation*}
For ease of notation we assume throughout that \( \Lambda_j(\blank \mid x) \) is
continuous for all \( x \) and \( j \). %ANDERS: need a note here which says that this is not a limitation?
We denote by \(S\) the conditional
event-free survival function:
\begin{equation}
  \label{eq:surv-def}
  S(t \mid x)=\exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)\right\}.
\end{equation}
% ANDERS: I prefer to have the \tau explicit
Let \( \mathcal{M}_{\tau}\) denote the space of all conditional cumulative hazard
functions on \( [0,\tau] \times\mathcal{X}\). Any distribution
\( Q \in \mathcal{Q} \) can be characterised by
\begin{equation*}
  \label{eq:parametrizeQ}
  \begin{split}
    Q(\diff t,j,\diff x)&= \left\{S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=1\}}}\\
                        &\quad\left\{S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{1_{\{j=2\}}},
  \end{split}
\end{equation*}
where \(\Lambda_{j} \in \mathcal{M}_{\tau}\) for \(j=1,2\) and \(H\) is the marginal
distribution of the covariates.

We consider the right-censored setting in which we observe data
\(O = (\tilde{T},\tilde D, X)\), where $\tilde T = \min(T,C)$ for a
right-censoring time \(C\), $\Delta = \1{\{T \leq C\}}$, and
\(\tilde D=\Delta D\). Let \(\mathcal{P}\) denote a set of probability
measures on the sample space
\(\mathcal{O} = [0, \tau] \times \{0, 1, 2\} \times \mathcal{X}\) such
that \(O \sim P \) for some unknown \(P\in \mathcal{P}\). We assume
that the event times and the censoring times are conditionally
independent given covariates, \( T \independent C \mid X \). This
implies that any distribution \( P \in \mathcal{P} \) is characterised
by a distribution \( Q \in \mathcal{Q} \) and a conditional cumulative
hazard function for \( C \) given \( X \)
\citep[c.f.,][]{begun1983information,gill1997coarsening}. We use
\(\Gamma\in\mathcal M\) to denote the conditional cumulative hazard
function for censoring. % ANDERS: can we not allow Gamma to be discrete?
For notational convenience, we assume that \( \Gamma(\blank \mid x) \) is
continuous for all \( x \), and let
\(G(t \mid x)=\exp\left\{-\Gamma(t \mid x)\right\}\) denote the
survival function of the conditional censoring distribution. The distribution 
is characterized by
\begin{equation}\label{eq:parametrizeP}
  \begin{split}
    P(\diff t, j, \diff x) &= \left\{G(t- \mid x)S(t- \mid x)\Lambda_1(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=1\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Lambda_2(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=2\}}}}\\
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}\\
                           &=\left\{G(t- \mid x)Q(\diff t,j,\diff x)\right\}^{\1{{\{j\ne 0\}}}}\\    
                           &\quad\left\{G(t- \mid x)S(t- \mid x)\Gamma(\diff t \mid x)H(\diff x)\right\}^{\1{{\{j=0\}}}}.
  \end{split}
\end{equation}
% ANDERS: H is missing in the model?
Hence, we may write
\( \mathcal{P} = \{ P_{Q, \Gamma} : Q \in \mathcal{Q}, \Gamma \in
\mathcal{G} \} \) for some \( \mathcal{G} \subset \mathcal{M}_{\tau} \). We
also have
\begin{equation*}
P(\tilde T>t \mid X=x) = S(t \mid x)G(t \mid x) = \exp\left\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x)-\Gamma(t \mid x) \right\}.
\end{equation*}
We further assume that there exists \(\kappa<\infty\) such that
\(\Lambda_{j}(\tau- \mid x)<\kappa \), for \(j\in\{1,2\}\), and
\(\Gamma(\tau- \mid x)<\kappa\) for almost all \(x\in\mathcal X\). Note that this
implies that \(G(\tau- \mid x)\) is bounded away from zero for almost all \(x\in\mathcal X\).
Under these assumptions, the conditional cumulative hazard functions
\(\Lambda_{j}\) and \(\Gamma\) can be identified from \(P\) by
\begin{align}
  \Lambda_{j}(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=j \mid X=x )}{P(\tilde T \geq s \mid X=x )}, \label{eq:lambdaj}\\
  \Gamma(t \mid x) &= \int_0^t\frac{  P(\tilde T \in \diff s, \tilde D=0 \mid X=x )}{P(\tilde T \geq s \mid X=x )}\label{eq:gamma}.
\end{align}
Thus, we can consider $\Lambda_j$ and \(\Gamma\) as operators which map from
\( \mathcal{P} \) to \(\mathcal M\).

\section{The concept of super learning}
\label{sec:super-learning}

In survival analysis, a super learner can be used to estimate a
parameter $\Psi$ which can be identified from the observed data
distribution \(P\in\mathcal P\). In this section, to introduce the
discrete super learner and the oracle learner, we consider estimation of the
function-valued parameter \(\Psi:\mathcal P\to\mathcal M\), given by
\(\Psi(P)=\Lambda_j\). This parameter is identified via equation
\eqref{eq:lambdaj} on \([0,\tau]\).

As input to the super learner we need a data set
\( \data_n=\{O_i\}_{i=1}^n \) of i.i.d.\ observations from \( P \in \mathcal{P} \) and a finite collection of candidate
learners $\mathcal{A}$. Each learner \(a \in \mathcal{A}\) is a map
\( a \colon \mathcal{O}^n \rightarrow \mathcal{M}_{\tau}\) which takes a data
set as input and returns an estimate $a(\data_n) \in \mathcal{M}_{\tau}$ of
$\Lambda_{j}$.
% Formally, the domain of $a$ depends on \(n\) but we suppress this in the notation.
In what follows, we use the short-hand notation
\(P[f] = \int f(o) P(\diff o) \). A super learner evaluates the
performance of \(a \in \mathcal{A}\) with a loss function
\(L\colon \mathcal{M}_{\tau} \times \mathcal{O} \rightarrow \R_+\) and
estimates the expected loss \(P[L(a(\data_n), \blank)]\) using
cross-validation. Specifically, the expected loss of $a\in\mathcal A$
is estimated by splitting the data set $\data_n$ into $K$ disjoint
approximately equally sized subsets
\(\data_n^1, \data_n^2, \dots, \data_n^K \) and then calculating the
cross-validated loss
\begin{equation*}
  % \label{eq:cv-risk-est}
  \hat{R}_n(a; L) =
  \frac{1}{K}\sum_{k=1}^{K}
  % \empmeas^k{[L {(a{ (\data_n^{-k})} , \blank) }]},
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  L
  {
    \left(
      a{ (\data_n^{-k})}
      , O_i
    \right)
  },
  \quad \text{with} \quad
  \data_n^{-k} = \data_n \setminus \data_n^{k}.
\end{equation*}
% where \( \empmeas^{k} \) is the empirical measure of \( \data_n^{k} \). 
The subset \(\data_n^{-k}\) is referred to as the \(k\)'th training
sample, while \(\data_n^{k}\) is referred to as the \(k\)'th test or
hold-out sample.
The discrete super learner is defined as
\begin{equation*}
\hat{a}_n = \argmin_{a\in\mathcal A}\hat{R}_n(a; L).
\end{equation*}
% The final estimator of \(\Psi(P)=\Lambda_j\) is then the selected
% learner applied to the full data set, i.e., \(\hat{a}_n(\data_n)\).
The oracle learner is defined as the learner that minimises the
average loss according to the data-generating distribution \( P \),
i.e.,
\begin{equation*}
  \tilde{a}_n =
  \argmin_{a \in \mathcal{A}}
  \tilde{R}_n(a; L),
  \quad \text{with} \quad 
  \tilde{R}_n(a; L)=
  \frac{1}{K}\sum_{k=1}^{K} 
  P{
    \left[
      L
      {
        \left(
          a{ (\data_n^{-k})}
          , \blank
        \right)
      }
    \right]}
  .
\end{equation*}
Note that both the discrete super learner and the oracle learner
depend on the library of learners and on the number of folds \(K\),
and that the oracle learner is a function of the data and the unknown
data-generating distribution. However, these dependencies are
suppressed in the notation.

@@

Machine learning based on right-censored data commonly use the
negative partial log-likelihood as loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
It seems however that this loss function is not perfectly suited for
super learning, because it cannot be applied to important survival
learners: The Kaplan-Meier estimator, the random survival forest, and
the semi-parametric Cox regression model provide cumulative hazard
functions that are piece-wise constant in the time argument, and hence
assign zero probability to event times not observed in the training
data. This implies that when event times are observed in continuous
time, any of these learners will almost surely have infinite loss in
an independent test set according to the negative partial
log-likelihood loss.
% ANDERS: would be good to have a formula for a the negative partial log-likelihood loss showing that the result is infitity

When a proportional hazards model is assumed, the baseline hazard
function can be profiled out of the likelihood to give a new partial
log-likelihood loss \citep{cox1972regression}, which has been
suggested as a loss function for super learning
\citep{golmakani2020super,verweij1993cross}. While this allows the
library of learners to include Cox' proportional hazard models, the
drawback is that the library is now restricted to include only these
models. % ANDERS: can the library really include multiple proportional hazards models? only one of them can be correct ...
% also, would be good to merge this paragraph with the previous somehow

@@

An attempt to avoid a pre-specified censoring model was proposed
independently by \cite{han2021inverse} and
\cite{westling2021inference}. In both articles, the authors suggest to
iterate between learning of \( \Lambda \) and learning of $\Gamma$
using IPCW loss functions. However, no general theoretical guarantees
seem to exist for this procedure, and it has not yet been extended to
the situation with competing risks.

@@

The state learner requires three libraries of learners,
\(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \), where
\(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain learners of the
conditional cause-specific cumulative hazard functions \(\Lambda_1\)
and \( \Lambda_2\), respectively, and \(\mathcal{B}\) contains
learners of the conditional cumulative hazard function of the
censoring distribution. % We further

@@

$\mathcal{F}$ of learners

@@

Based on a split of a data set \(\data_n\) into $K$ disjoint
approximately equally sized subsets (see Section \ref{sec:super-learning}), each learner

@@

\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \) is
evaluated using the cross-validated loss,

@@

and the discrete state learner is given by
